A Theory for Migration of Information Systems

### 2022-09-03 2022 Migration/articleMigration.tex
tussentijdse push

\section{Basics}
\label{sct:Basics}
	To define migration, let us first introduce the notion of dataset.
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
	To abstract away from a particular database implementation,
	we represent data elements as triples.
	This makes our theory valid for any kind of database that triple stores can represent,
	such as SQL databases, graph databases and other no-SQL databases

@@

\subsection{Triples}
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	Triples are used to relate two atoms and a relation.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	The practical meaning has no consequences in the formal world.
	The set of triples is called $\triples$.

\subsection{Relations}
	Relations are used in datasets to store data.
	In this paper relations are represented by variables $r$, $s$, and $d$.

	Every relation $r$ has a set of pairs, which we call the population of $r$:
\begin{equation}
	\pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
\end{equation}
	Every relation has a name, a source concept, and a target concept.
	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
	% To disassemble a relation in its name, source and target, there exist three functions:
% \[\begin{array}[3]{rcl}
	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
	% \id{src}&:&\rels\rightarrow\concepts\\
	% \id{tgt}&:&\rels\rightarrow\concepts
% \end{array}\]
	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
	% \(\src{\declare{nm}{A}{B}}=A\), and
	% \(\tgt{\declare{nm}{A}{B}}=B\).

	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
	The set of relations is called $\rels$.

@@

\subsection{Datasets}
	We can now define the notion of dataset.
	Let	$\concepts$ be a set of concept symbols,
	$\rels$ be a set of relations,
	$\atoms$ be a set of atoms,
	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
	% let $\roles$ be a set of roles,
	and $\rules$ be a set of constraints.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\rels,\atoms,\inst,\triples,\rules\ra$ with
	\[\begin{array}[3]{rcl}
		% \id{maintain}&:&\roles\times\rules\\
		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset
	\end{array}\]
\end{definition}

>>>>

\section{Basics}
\label{sct:Basics}
	To define migration, we must introduce the notion of dataset because migration aims to preserve data whilst changing the software that manipulates data.
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
	To abstract away from a particular database implementation,
	we represent data elements as triples.
	This makes our theory valid for any kind of database that triples can represent,
	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases

@@

\subsection{Triples}
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	Triples are used to relate two atoms and a relation.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	This ``meaning from practice'' has no consequences in the formal world.
	The set of triples is called $\triples$.

\subsection{Relations}
	Relations in datasets are used to store data.
	In this paper relations are represented by variables $r$, $s$, and $d$.

	Every relation $r$ contains a set of pairs, which we call the population of $r$:
\begin{equation}
	\pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
\end{equation}
	Every relation has a name, a source concept, and a target concept.
	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
	% To disassemble a relation in its name, source and target, there exist three functions:
% \[\begin{array}[3]{rcl}
	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
	% \id{src}&:&\rels\rightarrow\concepts\\
	% \id{tgt}&:&\rels\rightarrow\concepts
% \end{array}\]
	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
	% \(\src{\declare{nm}{A}{B}}=A\), and
	% \(\tgt{\declare{nm}{A}{B}}=B\).

	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	It contains all pairs that can be made of elements of $A$ and $B$:
\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A, \}\]
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
	The set of relations is called $\rels$.

@@

\subsection{Datasets}
	We can now define the notion of dataset.
	Let	$\concepts$ be a set of concept symbols,
	$\rels$ be a set of relations,
	$\atoms$ be a set of atoms,
	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
	% let $\roles$ be a set of roles,
	and $\rules$ be a set of constraints.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\isa,\rels,\atoms,\inst,\triples,\rules\ra$ with
	\[\begin{array}[3]{rcl}
		% \id{maintain}&:&\roles\times\rules\\
		\isa\text{\ is a partial order}\\
		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset
	\end{array}\]
\end{definition}

\subsection{Motivations}
	We have already motivated the use of triples in a dataset:
	It is a generic way of representing structured data, which maps to most (if not all) existing database technologies.
	Changes in the data means changing the set of triples, adding or removing triples from the set.
	That is not migration; that is regular operational behaviour.
	If the software changes, this means that the concepts, relations, and constraints change as well.
	However, a dataset can contain atoms that do not partake in any triple.
	That is why we need a set of atoms, $\atoms$ too.

	Atoms serve as data elements. Changes in the dataset are reflected in $\atoms$.
	Concepts serve to classify atoms in a partial order (the classification) so we can implement a typing regime on the dataset.
	This type system ensures that the dataset can be interpreted as a meaningful database.
	As a consequence, the relations $\isa$ and $\inst$

	The choice of elements in the tuple that represents a dataset is determined by the dynamics of a dataset.
	Anything that can change needs to be in that tuple.
	It is not only the data that changes. Rules can change, relations and concepts can change, the taxonomy can change,
	rules can change.
	In migrations we want to cope with any conceivable change,
	ensuring that the dataset satisfies its defining properties (from definition~\ref{def:dataset}).
	The migration must also ensure that the population remains intact as much as possible.
